---
title: "Preliminary Analysis for Appalachian Prison Water Contamination Exposure Project"
author: 
  - Kyle Onda, Center for Geospatial Solutions
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    title-block-banner: true
    toc: true
    toc-location: left
    toc-depth: 3
    html-math-method: katex
    css: styles.css
    code-fold: true
    code-summary: "show the code"
    code-overflow: wrap
    code-tools: true
    number-sections: true
    theme: 
      dark: slate
      light: flatly
editor: visual
---

## Introduction 

The purpose of this document is to demonstrate some data integration for analyzing water contaminant exposures in Appalachi

## Setup

Here we import necessary libraries.

```{r}
library(tidyverse)
library(sf)
library(mapview)
```

## Map main features of interest

### Get Geospatial Data

First we get the boundaries of the Appalachian Regional Commission (ARC) and its constituent counties

```{r}
arc <- sf::read_sf("https://services.arcgis.com/nkunl3y8FDxPkXDl/ArcGIS/rest/services/arc_boundary/FeatureServer/0/query?where=1%3D1&objectIds=&time=&geometry=&geometryType=esriGeometryEnvelope&inSR=&spatialRel=esriSpatialRelIntersects&resultType=none&distance=0.0&units=esriSRUnit_Meter&relationParam=&returnGeodetic=false&outFields=*&returnGeometry=true&returnCentroid=false&featureEncoding=esriDefault&multipatchOption=xyFootprint&maxAllowableOffset=&geometryPrecision=&outSR=&defaultSR=&datumTransformation=&applyVCSProjection=false&returnIdsOnly=false&returnUniqueIdsOnly=false&returnCountOnly=false&returnExtentOnly=false&returnQueryGeometry=false&returnDistinctValues=false&cacheHint=false&orderByFields=&groupByFieldsForStatistics=&outStatistics=&having=&resultOffset=&resultRecordCount=&returnZ=false&returnM=false&returnExceededLimitFeatures=true&quantizationParameters=&sqlFormat=none&f=pgeojson&token=")

counties <- sf::read_sf("https://reference.geoconnex.us/collections/counties/items?limit=4000")
states <- sf::read_sf("https://reference.geoconnex.us/collections/states/items?limit=4000")
counties_ct <- st_centroid(counties)
counties_ct <- counties_ct[arc,]
counties <- counties[which(counties$uri %in% counties_ct$uri),]
states <- states %>% filter(STATEFP %in% counties$STATEFP)
```

Now we get relevant public water systems from the Internet of Water geoconnex.us system

```{r}
download.file(url = "https://www.hydroshare.org/resource/3295a17b4cc24d34bd6a5c5aaf753c50/data/contents/ref_pws.gpkg" ,destfile="data/pws.gpkg")

pws_all <- read_sf("data/pws.gpkg") %>% st_make_valid()
pws <- pws_all %>% 
  st_intersects(arc) %>% lengths()

pws <- pws_all[which(pws==1),]

pws <- pws %>%
  select(pwsid,
         uri,
         pws_name,
         tier,
         primary_water_source,
         sdwis_link
         ) %>%
  mutate(sdwis_link =
           paste0('<a href="',
            sdwis_link,
            '">SDWIS link</a>')
  )
```

Now we get prisons in the region

```{r}
prisons_all <- sf::read_sf("https://opendata.arcgis.com/api/v3/datasets/2d6109d4127d458eaf0958e4c5296b67_0/downloads/data?format=geojson&spatialRefId=4326&where=1%3D1") 

prisons <- prisons_all %>% 
  st_intersects(arc) %>% lengths()

prisons <- prisons_all[which(prisons==1),]
prisons_ct <- st_centroid(prisons)
## filter to fed/state only
## visualize Status


```

### Make a map

```{r}


  mapview(arc,
        alpha.regions=0,
        color="black",
        col.regions="black",
        lwd=3,
        layer.name = "ARC Boundary", canvas=TRUE) +
  mapview(states,
          alpha.regions=0.5,
          col.regions="grey",
          lwd=2,
          layer.name="States", canvas=TRUE) +
   mapview(counties,
          alpha.regions=0.5,
          col.regions="grey",
          lwd=1,
          layer.name="Counties", canvas=TRUE) +
    mapview(pws,
          zcol="tier", canvas=TRUE) +
    mapview(prisons_ct, canvas=TRUE, cex=2) +
    mapview(prisons,canvas=TRUE, legend=FALSE)
```
